Distortion-corrected kernel density estimate on Riemannian manifolds

### 2022-07-18 paper/paper.tex
Sec 2.1-2.3 rewrite

For multivariate data, variable kernel density estimator with adaptive bandwidth matrix at each sample point can be used to capture the local structure of the data. \textcite{Pelletier2005-vu} generalize the multivariate kernel density esitmator to Riemnnian manifolds. However, it is not applicable when the underlying manifold is not known or when the true manifold is mapped in a much higher dimensional space. Therefore, we propose a distortion-corrected kernel density estimator as in Equation \eqref{eq:denestimator0} based on the low-dimensional embedding, \(\pmb{y}_i \in \mathbb{R}^d, i = 1, 2, \dots, N\), of the high-dimensional data with \(N\) data points,
\begin{equation}
\label{eq:denestimator0}
\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \frac{|\pmb{H}(\pmb{p})|}{|\pmb{H}(\pmb{y}_i)|} K\bigg( \frac{(\pmb{p} - \pmb{y}_i)^\prime H^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
\end{equation}
where \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\) is the pointwise adaptive Riemannian matrix of dimension \(d \times d\).
The proposed estimator could be used as an estimate of the density of the true manifold, with the local distortion at each point corrected by volume density functions. These densities could further be used for unsupervised tasks including anomaly detection when outliers are defined as the points with the lowest densities.

@@

For the exponential map \(\exp_p: T_pM \rightarrow M\) induced by pullback of a volume form \(\exp_p^*\textit{Vol}\) on \(T_pM\), \(\theta_p\) is its density with respect to the Lebesgue measure of the Euclidean structure on \(T_pM\), which gives
\[
d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du,
\]
in normal coordinates \autocite{Le_Brigant2019-lj}. The volume density function is motivated by how a measure \(du\) changes under the change of coordinates. The volume form is variant to the exponential map except that the manifold is flat. Therefore, a multiplicative corrected term called the volume density function has to be added in the estimator so that it will integral to 1. It is closely related to the integration on manifolds and plays an important part in studying the density of manifolds.
\textcite{Pelletier2005-vu} mentions that the volume density function \(\theta_q(p)\) defined for \(p\) in a neighborhood of \(q\) on the manifold, with the geodesic normal coordinates at \(q\), is equal to the determinant of the Riemannian metric \(g\) expressed in the coordinates at the logarithm map of q, \(\exp_q^{-1}(p)\).
Note that this is true only when using the geodesic normal coordinates. If the mapping function is not exponential/logarithm map, the volume density function is calculated otherwise. Therefore, the Riemannian manifold density estimator in \textcite{Pelletier2005-vu} only works when the true manifold \(M\) is known.

@@

Motivated by the estimator in \eqref{eq:denriem}, we restate our proposed distortion-corrected kernel density estimator (DC-KDE) of \(f\), same as \eqref{eq:denestimator0}, based on the manifold embedding \(\pmb{y}_i, i = 1, 2, \dots, N\),
\begin{equation}
\label{eq:denestimator}
\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \frac{|\pmb{H}(\pmb{p})|}{|\pmb{H}(\pmb{y}_i)|} K\bigg( \frac{(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
\end{equation}
where \(r\) is a global bandwidth scalar, \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\) is the pointwise variable Riemannian matrix and it is positive definite with dimension \(d\times d\), \(\frac{|\pmb{H}(\pmb{p})|}{|\pmb{H}(\pmb{y}_i)|}\) is the estimated volume density function when using the coordinates from any manifold learning mapping, \((\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)\) is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric \(\pmb{H}^{-1}(\pmb{y}_i)\). Then the only problem left is the estimation of the Riemannian metric at each embedding point, discussed in \autoref{riemmetric}.

By the definition of the volume density function in Equation \eqref{eq:volden}, we could calculate how the volume change under the change of coordinates. When using the logarithm map, the coordinates of points on the manifold are transformed into their coordinates in the tangent space, i.e.~geodesic normal coordinates. In this case, \textcite{Pelletier2005-vu} gives the volume density function from the manifold to the tangent space which is equal to the determinant of the Riemannian metric expressed in the geodesic normal coordinates at the logarithm map, \(|\det g(\exp^{-1}_p(q)) |\).
In order to estimate the density of the true manifold, we first map the high-dimensional data to a low-dimensional embedding, then we apply a kernel density estimator to the embedding. Since the logarithm map of the manifold is also unknown, the volume density function in \textcite{Pelletier2005-vu} is no longer valid. Instead, we need a volume density function independent of the chart.

For \(p \in M\), denote an open ball centered at \(p\) with radius \(r\) as \(B(p,r)\) and the open neighborhood of \(0_p \in T_pM\) can be denoted as \(B(0_p,r)=\exp_p^{-1}(B(p,r))\). Consider the Riemannian metrics \(g^{\prime}\) and \(g^{\prime \prime}\) in \(B(0_p,r)\), with \(g^{\prime} = \exp^{\ast}_p(g)\) being the pullback of \(g\) induced by the exponential map and \(g^{\prime \prime}\) being the canonical metric induced by \(g_p\) in \(B(0_p,r)\).
Let \(u \in B(0_p,r)\) be a tangent vector and for any given chart \((\overline{U}, \overline{\psi})\) of \(B(0_p,r)\), we have \(u \in \overline{U}\).
The volumes of the parallelepiped spanned by the basis \(\{(\partial / \partial\overline{\psi}_1|_u), \dots, (\partial / \partial\overline{\psi}_d|_u)\}\) are given by \(| \det g^{\prime}((\partial / \partial\overline{\psi}_i|_u), (\partial / \partial\overline{\psi}_j|_u)) |^{1/2}\) with respect to \(g^{\prime}\), and \(| \det g^{\prime \prime}((\partial / \partial\overline{\psi}_i|_u), (\partial / \partial\overline{\psi}_j|_u)) |^{1/2}\) with respect to \(g^{\prime \prime}\). The quotient of these two volumes is then independent of the chart \((\overline{U}, \overline{\psi})\) \autocite{Henry2009-om}.
Therefore, given a point \(q \in B(p,r)\) and a tangent vector \(u = \exp^{-1}_p(q) \in B(0_p,r)\), for any chart \((\overline{U}, \overline{\psi})\) of \(B(0_p,r)\) that contains \(u\), \textcite{Henry2009-om} defines the volume density function on the Riemannian manifold \((M,g)\) as
\[
\theta_p(q) = \frac{| \det g^{\prime}((\partial / \partial\overline{\psi}_i|_u), (\partial / \partial\overline{\psi}_j|_u)) |^{1/2} }{ | \det g^{\prime \prime}((\partial / \partial\overline{\psi}_i|_u), (\partial / \partial\overline{\psi}_j|_u)) |^{1/2} } \\ .
\]
The volume density function is only defined for points satisfying \(d_g(p,q) < inj_gM\) so that there exists an open ball \(B(p,r) \in T_pM\). And \(\theta_p(q)=1\) for all \(p,q \in \mathbb{R}^d\) when the manifold \(M\) is \(\mathbb{R}^d\) with the canonical metric.

Now we consider the case with the manifold learning embedding. Consider the chart \((U, \psi)\) induced by an orthonormal basis of the tangent space of the embedding \(T_{\pmb{y}_i}E\) and a normal neighborhood of \(\pmb{y}_i\). Given a query point \(\pmb{p} \in B(\pmb{y}_i,r)\), the volume of the parallelepiped is spanned by the same basis.\\
Then the volume density function is given by
\[
\theta_{\pmb{y}_i}(\pmb{p}) = \frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}},
\]
where \(\pmb{H}\) is the estimated Riemannian metric for the unknown manifold.
Again, the volume density function is not defined for all \(\pmb{p}\) but for points within the geodesic ball satisfying \(\pmb{p} \in B(\pmb{y}_i,r)\), which is related to the support of \(K\) being \([0,1]\) in (v) of \eqref{eq:kernelcondition}.

@@

As pointed out by \textcite{Perrault-Joncas2013-pq}, the naive distance \(\|p-q\|\) underestimates \(d_g\), and the graph shortest distance \(d_G\) overestimates \(d_g\) by the triangle inequality. By using only the first-order correction in \(\hat{d}_g\), our estimator is also affected by the estimated Riemannian metric. In order to correct for the estimation error in \(\pmb{H}(\pmb{y}_i)\), we further divide the density estimator in \eqref{eq:denestimator} by a constant term \(C\) so that the integral of the kernel is equal to 1. The constant term in \(1\)-d is given by the \emph{area under the curve (AUC)}, and can be named as the \emph{volume between the surface} in higher dimension. We use the volume estimator in Section 6.4 of \textcite{Perrault-Joncas2013-pq} to sum over all embedding points \(\pmb{y}_i\) and get
\[
C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i),
\]
where the distortion is corrected for the tangent plane of the embedding points.

>>>>

For multivariate data, variable kernel density estimator with adaptive bandwidth matrix at each sample point can be used to capture the local structure of the data. \textcite{Pelletier2005-vu} generalize the multivariate kernel density esitmator to Riemnnian manifolds. However, it is not applicable when the underlying manifold is not known or when the true manifold is mapped in a much higher dimensional space. Therefore, we propose a distortion-corrected kernel density estimator as in Equation \eqref{eq:denestimator0} based on the low-dimensional embedding, \(\pmb{y}_i \in \mathbb{R}^d, i = 1, 2, \dots, N\), of the high-dimensional data with \(N\) data points,
\begin{equation}
\label{eq:denestimator0}
\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}} K\bigg( \frac{ \sqrt{(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)} }{r} \bigg),
\end{equation}
where \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\) is the pointwise adaptive Riemannian matrix of dimension \(d \times d\).
The proposed estimator could be used as an estimate of the density of the true manifold, with the local distortion at each point corrected by volume density functions. These densities could further be used for unsupervised tasks including anomaly detection when outliers are defined as the points with the lowest densities.

@@

For the exponential map \(\exp_p: T_pM \rightarrow M\) induced by pullback of a volume form \(\exp_p^*\textit{Vol}\) on \(T_pM\), \(\theta_p\) is its density with respect to the Lebesgue measure of the Euclidean structure on \(T_pM\), which gives
\[
d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du,
\]
in normal coordinates \autocite{Le_Brigant2019-lj}. The volume density function is motivated by how a measure \(du\) changes under the change of coordinates. The volume form is variant to the exponential map except that the manifold is flat. Therefore, a multiplicative corrected term called the volume density function has to be added in the estimator so that it will integral to 1. It is closely related to the integration on manifolds and plays an important part in studying the density of manifolds.
\textcite{Pelletier2005-vu} mentions that the volume density function \(\theta_q(p)\) defined for \(p\) in a neighborhood of \(q\) on the manifold, with the geodesic normal coordinates at \(q\), is equal to the determinant of the Riemannian metric \(g\) expressed in the coordinates at the logarithm map of q, \(\exp_q^{-1}(p)\).
Note that this is true only when using the geodesic normal coordinates. If the mapping function is not exponential/logarithm map, the volume density function is calculated otherwise. Therefore, the Riemannian manifold density estimator in \textcite{Pelletier2005-vu} only works when the true manifold \(M\) is known.
Further, the fixed bandwidth assumes that the smoothing is the same across the entire manifold and there is no local distortion at each sample point, which needs to be corrected using an adaptive bandwidth matrix when distortion exists.

@@

Motivated by the estimator in \eqref{eq:denriem}, we restate below our proposed distortion-corrected kernel density estimator (DC-KDE) of \(f\), same as \eqref{eq:denestimator0}, based on the manifold embedding \(\pmb{y}_i, i = 1, 2, \dots, N\),
\begin{equation}
\label{eq:denestimator}
\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}} K\bigg( \frac{ \sqrt{(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)} }{r} \bigg),
\end{equation}
where \(r\) is a global bandwidth scalar, \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\), is the pointwise variable Riemannian matrix and it is positive definite with dimension \(d\times d\), \(\frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}}\) is the estimated volume density function when using the coordinates from any manifold learning mapping, \((\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)\) is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric \(\pmb{H}(\pmb{y}_i)\), and \(C\) is a correction term for the volume between the spaces so that the integral is 1. We will now state below how to calculate each term in the estimator. Then the only problem left is the estimation of the Riemannian metric at each embedding point, to be discussed in \autoref{riemmetric}.

By the definition of the volume density function in Equation \eqref{eq:volden}, we could calculate how the volume change under the change of coordinates. When using the logarithm map, the coordinates of points on the manifold are transformed into their coordinates in the tangent space, i.e.~geodesic normal coordinates. In this case, \textcite{Pelletier2005-vu} gives the volume density function from the manifold to the tangent space, which is equal to the determinant of the Riemannian metric expressed in the geodesic normal coordinates at the logarithm map, \(|\det g(\exp^{-1}_p(q)) |\).
In order to estimate the density of the true manifold, we first map the high-dimensional data to a low-dimensional embedding, then we apply a kernel density estimator to the embedding. Since the logarithm map of the manifold is also unknown, the volume density function in \textcite{Pelletier2005-vu} is no longer valid. Instead, we need a volume density function independent of the chart.

For \(p \in M\), denote an open ball centered at \(p\) with radius \(r\) as \(B(p,r)\) and the open neighborhood of \(0_p \in T_pM\) can be denoted as \(B(0_p,r)=\exp_p^{-1}(B(p,r))\). Consider the Riemannian metrics \(g^{\prime}\) and \(g^{\prime \prime}\) in \(B(0_p,r)\), with \(g^{\prime} = \exp^{\ast}_p(g)\) being the pullback of \(g\) induced by the exponential map and \(g^{\prime \prime}\) being the canonical metric induced by \(g_p\) in \(B(0_p,r)\).
Let \(u \in B(0_p,r)\) be a tangent vector and for any given chart \((\bar{U}, \bar{\psi})\) of \(B(0_p,r)\), we have \(u \in \bar{U}\).
The volumes of the parallelepiped spanned by the basis \(\{(\partial / \partial\bar{\psi}_1|_u), \dots, (\partial / \partial\bar{\psi}_d|_u)\}\) are given by \(| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2}\) with respect to \(g^{\prime}\), and \(| \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2}\) with respect to \(g^{\prime \prime}\). The quotient of these two volumes is then independent of the chart \((\bar{U}, \bar{\psi})\) \autocite{Henry2009-om}.
Therefore, given a point \(q \in B(p,r)\) and a tangent vector \(u = \exp^{-1}_p(q) \in B(0_p,r)\), for any chart \((\bar{U}, \bar{\psi})\) of \(B(0_p,r)\) that contains \(u\), \textcite{Henry2009-om} defines the volume density function on the Riemannian manifold \((M,g)\) as
\[
\theta_p(q) = \frac{| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} }{ | \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} } \\ .
\]
The volume density function is only defined for points satisfying \(d_g(p,q) < inj_gM\) so that there exists an open ball \(B(p,r) \in T_pM\). And \(\theta_p(q)=1\) for all \(p,q \in \mathbb{R}^d\) when the manifold \(M\) is \(\mathbb{R}^d\) with the canonical metric.

Now we consider the case with the manifold learning embedding. Consider the chart \((U, \psi)\) induced by an orthonormal basis of the tangent space of the embedding \(T_{\pmb{y}_i}E\) and a normal neighborhood of \(\pmb{y}_i\). Given a query point \(\pmb{p} \in B(\pmb{y}_i,r)\), the volume of the parallelepiped is spanned by the same basis.\\
Then the volume density function is given by
\[
\theta_{\pmb{y}_i}(\pmb{p}) = \frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}},
\]
where \(\pmb{H}\) is the estimated Riemannian metric for the unknown manifold.
Again, the volume density function is not defined for all \(\pmb{p}\) but for points within the geodesic ball satisfying \(\pmb{p} \in B(\pmb{y}_i,r)\), which is related to the support of \(K\) being \([0,1]\) in condition (v) of \eqref{eq:kernelcondition}.

@@

As pointed out by \textcite{Perrault-Joncas2013-pq}, the naive distance \(\|p-q\|\) underestimates \(d_g\), and the graph shortest distance \(d_G\) overestimates \(d_g\) by the triangle inequality. By using only the first-order correction in \(\hat{d}_g\), our estimator is also affected by the estimated Riemannian metric. In order to correct for the estimation error in \(\pmb{H}(\pmb{y}_i)\), we further divide the density estimator in \eqref{eq:denestimator} by a constant term \(C\) so that the integral of the kernel is equal to 1. The constant term in \(1\)-d is given by the \emph{area under the curve (AUC)}, and can be named as the \emph{volume between the surface (VBS)} in higher dimension. We use the volume estimator in Section 6.4 of \textcite{Perrault-Joncas2013-pq} to sum over all embedding points \(\pmb{y}_i\) and get
\[
C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i),
\]
where the distortion is corrected for the tangent plane of the embedding points.

Now we have all the components needed to estimate the density of manifold learning embedding in Equation \eqref{eq:denestimator} except for the Riemannian metric. The Riemannian metric is a key to correct the distortion from the manifold learning algorithms and we discuss its estimation in \autoref{riemmetric}.
